---
title: "Packages & Reproducibility: Install what you need, attach what you want"
author: Mike DeCrescenzo
description: "A comment on conditionally installing, attaching, and not attaching packages for reproducible R scripts"
date: '2018-05-30'
slug: reproducible-packages
categories: ["Methods"]
tags: ["Reproducibility", "R"]
comments: no
showcomments: yes
showpagemeta: yes
draft: false
---

*(Note: An earlier version of this post referred to "loading" packages when I really meant "attaching." Thanks to [Thomas Leeper](https://twitter.com/thosjleeper/status/1001859564113924096) for the clarification.)*

---

When we distribute R code (for publication/replication archives, on Github, through blog posts, etc), we like the code to run smoothly on someone else's machine. Packages present a nominal problem because different users have different packages installed on their computer. Ideally the script we are distributing should install dependencies without the redundancy of re-installing packages that a user already has installed. 

To solve this, many users conditionally install packages using `require()`; if a package fails to attach, it is installed. Supposing that we want to use a package called `pkg`...

```{r, include = FALSE}
# future chunks are not evaluated
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE)
```

```{r}
# if require() succeeds, package is attached
# if require() fails, package is installed
if (require("pkg") == FALSE) {
  install.packages("pkg")
}
```



We see often see this in (for example) package ReadMe files on Github. But for distributing bigger projects, the approach has two shortcomings.

1. Sometimes we require a package, but we don't want to attach it fully. We may only need it for a function or two, or we want to prevent [clashing function names](https://github.com/r-lib/conflicted) in projects that use many packages, and so we prefer to use the `pkg::function()` grammar instead of attaching every package with `library()`.
2. If we require several packages for a large project (like a replication archive for a published journal article), including code to install all packages ends up unsightly and repetitive (but functional, to be fair).

This post discusses a straightforward strategy for addressing this.


## Check for installed packages without attaching them

Rather than use `require()`, which has the side-effect of attaching an installed package, we can look for installed packages using the output from `installed.packages()`. In particular, the function should return an array whose rownames contain the names of our installed packages. 

```{r, eval = TRUE}
head(rownames(installed.packages()))
```

Using these rownames, we can check for installed packages without attaching those packages incidentally. Just create a character vector of required packages and cross-reference those names with the list of installed packages.

```{r, eval = TRUE}
c("ggplot2", "beepr", "some_other_package") %in% rownames(installed.packages())
```


## Install what we need, attach what we want

The structure laid out above can be augmented to install missing packages. Just use a little bit of indexing:

```{r}
# vector of requirements 
requires <- c("pkg_1", "pkg_2")

# evaluates to TRUE if not already installed
to_install <- (requires %in% rownames(installed.packages()) == FALSE)

# install missing packages
cloud_url <- "https://cloud.r-project.org/"
install.packages(requires[to_install], repos = cloud_url)
```

Because we never call `require()`, we can attach (or not attach) whichever packages we want with subsequent `library()` commands. (We could have also used `requireNamespace()` to check for packages.)



# Coda: on the use of `library()` vs `require()`

Thinking back to this [oft-cited post by Yihui Xie](https://yihui.name/en/2014/07/library-vs-require/), the takeaway was that we should only be using `require()` to...

1. conditionally install packages, or 
2. implement "bonus-features" that may enhance a package or function but that aren't strictly necessary

In reference to [1], I think we can (and should) flatly avoid `require()` in any situation where we need a package but don't want to attach it. For me, this is basically all of my research projects, and I think many other users will find themselves in a similar boat. As long as `require` attaches a package in the process of checking for it, `require` gives unintended and unnecessary side-effects.

In reference to [2], it's a little funny that we got to a point where we would only use a function called `require` for features that are strictly not required. It's quite the contradiction. In a parallel universe we might have named the function something squishier like `suppose()`. At any rate, the chief reason to use `require` in this case---it returns logical output---isn't necessary either, since `library()` has a `logical.return` argument that achieves the same objective.


# Thanks for reading

Feel free to [get in touch](https://www.twitter.com/mikedecr) with comments, links to related posts by others, packages that implement similar features, and so on. 




